---
title: "Fairness"
author: "Gizelle H"
date: "11/07/2020"
csl: american-medical-association.csl
output:
  bookdown::html_document2:
    number_sections: false
  bookdown::word_document2:
    number_sections: false    
  bookdown::pdf_document2:
    number_sections: false
    toc: false
    keep_tex: yes
link-citations: yes
bibliography: bibliography.bib
---

# Fairness in Trauma Prediction Models

## Introduction

Trauma continues to be a major contributor to the global burden of
disease - an estimated 520 million injuries required health
interventions in 2017 alone [@James2018]. Annually, nearly 4.5 million
people die due to trauma and over 90% of these deaths occur in low-
and middle-income countries [@Roth2018].

Trauma mortality prediction models aim to estimate the probability of
death in trauma patients and improve quality control and evaluation of
trauma systems [@Cook2014]. The most commonly used trauma mortality
prediction models are Trauma Mortality Prediction Model (TMPM), New/Injury
Severity Score (N/ISS) and the Revised Trauma Score (RTS).

Trauma databases include more men than women [@Haider2009; @Meng2019],
and most trauma severity scores were developed in overwhelmingly male
populations [@Akay2017; @Tohira2012]. The effect of this inherent bias
on the fairness (distribution of model errors) has not been previously
studied despite potentially affecting the quality of care provided to
women.

The study of fairness in trauma prediction models can help identify
and correct potential flaws in these models and improve trauma
outcomes by informing better treatment protocols. The aim of this
study is to assess and compare the fairness of various trauma
prediction models.


## Methods

### 1.1 Study design & data sources

Our study is a retrospective analysis of the anonymous multicenter
Towards Improved Trauma Care Outcomes (TITCO) in India cohort.

### 1.2 Key study dates

We conducted this analysis between [dates] based on data directly
observed or retrieved from patient records between July 2013 and
December 2015.

### 2.1 Study setting

The patient data was collected from four hospitals within urban,
Indian cities of Mumbai, New Delhi and Kolkata. This included two
tertiary care hospitals with dedicated trauma wards and one without,
while the fourth was a dedicated trauma center.

### 2.2 Eligibility criteria

The TITCO cohort includes all patients presenting with trauma, alive
at the time of arrival, and who were admitted to the participating
hospitals, regardless of age or gender. Those dying between arrival
and admission were also included. The exclusion criteria were subjects
dead on arrival and those presenting exclusively with peripheral
injuries.

### 3.1 Predicted outcome

The predicted outcome is death following trauma with or without
surgical intervention, from time of arrival to hospital to 30 days after.

### 3.2 Actions to blind outcome assessment

All data was collected based on standardized criteria regardless of
demographic details and outcomes, including all predefined predictors
as applicable.

### 4.1 Predictors and measurement

- AIS score
- ICD-10 coding
- SBP
- RR
- GCS
- Age (index)
- Number of injuries

### 4.2 Actions to blind predictor assessment

- Ensuring expert opinion in recording intraoperative diagnoses and assigning primary injury.
- Following standardized coding methods (ICD-10)
- Ensuring all coders receive similar standardized training for data collection. 
	

### 5. Sample size

```{r sample-size-calculation, echo = FALSE, include = FALSE}

library(pwr)

stats <- pwr.t.test(d = 0.2, power = 0.8, type = "two.sample", alternative = "two.sided")
stats.paired <- pwr.t.test(d = 0.2, power = 0.8, type = "paired", alternative = "two.sided")

```
The TITCO cohort includes 16,000 trauma patients admitted to the four
participating hospitals. We need at least `r stats$n` men and as many
women, coming to a minimum total sample size of `r stats$n*2`, to
detect an effect size of 0.2 (considered a small effect size
[@Cohen1988; @Lakens2013]) when comparing the mean absolute prediction
error between men and women for any model with 80% power and a
significance level of 5%.

### 6. Missing data
We will perform a complete case analysis, meaning that patients with
missing values in any predictor or outcome will be excluded.

### 7.1 Predictor analysis

Predictor value description, separately for men and women.

### 7.2 Type of model and predictor selection

The data collected in the TITCO cohort allows for the comparison of
the ISS, NISS, TMPM-ICD-10-CM, KTS, TRISS and ASCOT models. The ISS,
New ISS and TMPM are based on anatomic region of injury [@Haider2012]
and use the Abbreviated Injury Scale (AIS) which assigns scores based
on region and type of anatomic structure injured and the severity of
the injury.

Other models use a combination of parameters. The Kampala Trauma Score
(KTS) uses age, systolic blood pressure (SBP), respiratory rate (RR),
neurological status and number of injuries; Revised Trauma Score (RTS)
uses the Glasgow Coma Scale (GCS), SBP and RR [@Akay2017]; Trauma and
Injury Severity Score (TRISS) is a combination of RTS, ISS and age
index; while A Severity Characterization of Trauma (ASCOT) uses AIS,
RTS and age [@Champion1996]

### 7.3 Measures used to assess and compare model performance

The mean absolute prediction error for each specified model was
calculated and is presented separately for men and women. We also
calculated the absolute and relative difference in mean absolute
prediction error between men and women. We compare these differences
using independent t-tests. We also compared differences in mean absolute
prediction error between men and women across models in a pair wise
fashion using paired t-tests. A model with a smaller difference in
mean absolute prediction error was considered more fair.

# Results

```{r dataset, echo=FALSE, include=FALSE}
## Load required packages
library(rio)
library(dplyr)

## Import data
titco <- rio::import("titco-I-full-dataset-v1.csv")

## Keep only required variables
study.data <- titco %>%
select(age,sex,tran,doi,toi,doar,toar,doa,toa,ti,sbp_1,rr_1,gcs_t_1,gcs_e_1,gcs_v_1,gcs_m_1,dom_1,tom_1,sbp_2,rr_2,gcs_t_2,gcs_t_2,gcs_e_2,gcs_v_2,gcs_m_2,dom_2,tom_2,sbp_s,ct,fast,xray,died,dodd,todd,dama,head_and_neck,face,chest,abdomen_and_pelvic_contents,extremities,external,iss,niss,e_1_icd,e_2_icd,e_3_icd,e_4_icd,e_5_icd,e_6_icd,e_6_icd,e_7_icd,e_8_icd,e_9_icd,e_10_icd,e_11_icd,xray_1_icd,xray_2_icd,xray_3_icd,xray_4_icd,xray_5_icd,xray_6_icd,xray_7_icd,xray_8_icd,xray_9_icd,xray_10_icd,xray_11_icd,fast_1_icd,fast_2_icd,fast_3_icd,fast_4_icd,fast_5_icd,fast_6_icd,fast_7_icd,fast_8_icd,fast_9_icd,fast_10_icd,fast_11_icd,ct_1_icd,ct_2_icd,ct_3_icd,ct_4_icd,ct_5_icd,ct_6_icd,ct_7_icd,ct_8_icd,ct_9_icd,ct_10_icd,ct_11_icd,ct_12_icd,ct_13_icd,op_1_icd,op_2_icd,op_3_icd,op_4_icd,op_5_icd,op_6_icd,op_7_icd,op_8_icd,op_9_icd,op_10_icd,op_11_icd) %>% as_tibble()

## Now go ahead and create the study sample, the sample that includes
## only the patients with complete data on all predictors and the
## outcome. You need to decide how to handle missing data in ICD and
## AIS columns, as that data is not "really" missing. 

```

The complete TITCO cohort includes `r nrow(titco)` patients. 


## References
